---
title: "Causal Inference in High-Dimensional Observational Data"
subtitle: "A Comparative Simulation Study of TMLE and C-TMLE Estimators"
author: "Juliet Asantewaa Sarpong"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 9,
  fig.height = 5
)

library(here)
```

------------------------------------------------------------------------

# Introduction

This notebook presents a comparative simulation study of four causal inference estimators for the **Average Treatment Effect (ATE)** in observational data settings. The focus is on high-dimensional settings where the number of covariates is large relative to sample size — a common challenge in healthcare and genomic research.

## The Causal Inference Problem

In observational studies, we cannot randomly assign treatment. Patients who receive a treatment may systematically differ from those who don't, in ways that also affect the outcome. This is **confounding**, and naively comparing treatment groups produces biased causal estimates.

The **structural causal model** underlying this study assumes:

-   **W** — a vector of pre-treatment covariates (potential confounders)
-   **A** — binary treatment indicator (A = 1 treated, A = 0 control)
-   **Y** — continuous outcome

The **Average Treatment Effect** is defined as:

$$\text{ATE} = E[Y(1) - Y(0)]$$

where Y(a) denotes the potential outcome under treatment a. Under the standard causal assumptions (consistency, positivity, and conditional exchangeability), the ATE is identified from observed data as:

$$\text{ATE} = E_W[E[Y \mid A=1, W] - E[Y \mid A=0, W]]$$

## Estimators Compared

Four estimators are benchmarked:

| Estimator | Abbreviation | Key Feature |
|------------------------|------------------------|------------------------|
| Targeted Maximum Likelihood Estimation | TMLE | Doubly robust; semiparametrically efficient |
| Collaborative TMLE (Greedy) | C-TMLE Greedy | Greedy forward selection of propensity score covariates |
| Collaborative TMLE (Lasso) | C-TMLE1 | Lasso regularisation path for propensity score |
| Collaborative TMLE (Gradient) | C-TMLE0 | Gradient-based collaborative selection |

All estimators use **cross-validated LASSO/elastic net** (`glmnet`) for nuisance model estimation, making them suitable for high-dimensional settings. Standard errors are computed via the **efficient influence function**, providing valid inference without parametric assumptions.

------------------------------------------------------------------------

# Setup

## Load Scripts

```{r source_scripts}
# Source auxiliary scripts
source(here("scripts", "01_data_generating_processes.R"))
source(here("scripts", "02_utilities.R"))
source(here("scripts", "03_estimators.R"))
source(here("scripts", "04_estimators_highdim.R"))
```

## Load Packages

```{r load_packages}
# Data manipulation
library(tidyverse)
library(MASS)
library(survey)
library(boot)

# Visualisation
library(ggplot2)
library(skimr)

# Learning algorithms
library(glm2)
library(glmnet)
library(randomForest)
library(SuperLearner)
```

------------------------------------------------------------------------

# Simulation Design

## Data Generating Process (Study 3 — High-Dimensional Sparse)

The primary simulation uses a **high-dimensional sparse DGP** with p = 100 covariates drawn from a multivariate normal distribution with Toeplitz covariance structure (ρ = 0.5), mimicking correlated genomic or clinical covariates.

Key design choices:

-   **p = 100** covariates, but only **k = 10** are truly predictive of the outcome, and **k = 10** of the treatment — **sparse signal** in both mechanisms
-   Treatment A is binary, generated from a logistic model with sparse coefficients
-   Outcome Y is continuous: `Y = 2·A + W·γ + ε`, where ε \~ N(0,1)
-   **True ATE = 2** (known by construction, enabling bias evaluation)

```{r preview_data}
# Generate a single dataset for illustration
set.seed(42)
data_example <- sim3(n = 1000, p = 100, rho = 0.5, k = 10, k2 = 10)

# Preview structure
cat("Dataset dimensions:", nrow(data_example), "observations,", ncol(data_example), "variables\n")
cat("Treatment prevalence:", round(mean(data_example$A), 3), "\n")
cat("Outcome mean:", round(mean(data_example$Y), 3), "\n")
cat("Outcome SD:", round(sd(data_example$Y), 3), "\n")
```

```{r skim_data}
# Summary of key variables
data_example %>%
  dplyr::select(W1, W2, W3, A, Y) %>%
  skimr::skim()
```

------------------------------------------------------------------------

# Single Dataset Results

Before running the full Monte Carlo study, we verify each estimator on a single dataset to confirm correct implementation and reasonable point estimates.

## Define Estimator List

```{r estimator_list}
# Wrap estimators with true_value = 2 (known ATE for Study 3)
estimator_list <- list(
  "TMLE"          = function(data) estimate_TMLE_glmnet(data, true_value = 2),
  "Greedy C-TMLE" = function(data) estimate_GREEDY_CTMLE_glmnet(data, true_value = 2),
  "C-TMLE0"       = function(data) estimate_CTMLE0_glmnet(data, true_value = 2),
  "C-TMLE1"       = function(data) estimate_CTMLE1_glmnet(data, true_value = 2)
)
```

## Single Dataset Estimates

```{r single_results}
set.seed(123)
data_test <- sim3(n = 1000, p = 100)

# Evaluate all estimators on a single dataset
single_results <- evaluate_estimators(data_test, estimator_list)

single_results %>%
  dplyr::select(Algorithm, est_init = Initial.Est, est = Estimate, se = Standard.Error,
         ci_lower = CI_Lower, ci_upper = CI_Upper, bias = Bias) %>%
  knitr::kable(digits = 3, caption = "Single dataset results — True ATE = 2")
```

All four estimators should recover an estimate close to the true ATE = 2. Initial estimates (before the targeting step) and targeted estimates (after) are both shown, allowing us to see the impact of the TMLE fluctuation step.

------------------------------------------------------------------------

# Monte Carlo Simulation

## Running the Study

The Monte Carlo study runs **k = 100 replications** with **n = 5,000** observations per dataset. This sample size is deliberately large to allow the asymptotic properties of the estimators to manifest clearly.

> ⚠️ **Note:** The full Monte Carlo run is computationally intensive (\~30–60 minutes). Pre-computed results are loaded below. To re-run from scratch, uncomment the `MC_est()` call.

```{r monte_carlo, eval=FALSE}
# --- OPTIONAL: Re-run the full Monte Carlo study ---
# combined_results_ <- MC_est(
#   k          = 100,
#   sample_no  = 5000,
#   simulation = sim3,
#   est_list   = estimator_list
# )
# write.csv(combined_results_, "../results/ctmle_glmnet.csv", row.names = FALSE)
```

```{r load_results}
# Load pre-computed Monte Carlo results
combined_results_ <- read.csv("../results/ctmle_glmnet.csv")

cat("Loaded", nrow(combined_results_), "rows across", 
    n_distinct(combined_results_$Algorithm), "estimators\n")
```

## Aggregate Performance Metrics

```{r mc_summary}
true_value <- 2

res <- MC_res(combined_results_, true_value = true_value)

res %>%
  dplyr::select(Algorithm, Mean_Est, Mean_Bias, Mean_Var, Coverage, Mean_MSE) %>%
  knitr::kable(
    digits  = 3,
    caption = "Monte Carlo Summary — k=100 replications, n=5,000, True ATE = 2",
    col.names = c("Estimator", "Mean Estimate", "Mean Bias", 
                  "Mean Variance", "CI Coverage", "MSE")
  )
```

**Interpreting the metrics:**

-   **Mean Estimate** — should be close to 2.0; deviation indicates bias
-   **Mean Bias** — directly measures systematic error; closer to 0 is better
-   **Mean Variance** — average squared standard error across replications
-   **CI Coverage** — proportion of 95% CIs containing the true ATE; should be close to 0.95
-   **MSE** — combines bias and variance; primary summary of overall performance

------------------------------------------------------------------------

# Visualisation

## Distribution of ATE Estimates by Estimator

The density plots below show the sampling distribution of each estimator's ATE estimate across 100 Monte Carlo replications. The dashed vertical line marks the **true ATE = 2**.

```{r dist_per_estimator, fig.height=6}
hist_per_est(
  combined_results_,
  xlimits    = range(combined_results_$Estimate),
  true_value = true_value
) +
  coord_cartesian(xlim = c(1, 3)) +
  labs(caption = "Dashed line = True ATE (2.0). Tighter distributions indicate lower variance.")
```

A well-performing estimator will have its distribution **centred on 2.0** (low bias) and **narrow** (low variance). C-TMLE variants are expected to show reduced variance compared to standard TMLE in this high-dimensional sparse setting, due to their collaborative variable selection in the propensity score model.

## Overlaid Estimate Distributions

```{r dist_all, fig.height=5}
hist_all(
  combined_results_,
  xlimits    = range(combined_results_$Estimate),
  true_value = true_value
) +
  coord_cartesian(xlim = c(1, 3)) +
  labs(caption = "All four estimators overlaid. Dashed line = True ATE (2.0).")
```

## Confidence Interval Coverage

The plots below show the distribution of CI lower and upper bounds across replications. For a well-calibrated estimator, the true ATE (2.0) should fall within approximately 95% of the plotted intervals.

```{r ci_distribution, fig.height=6}
hist_per_est_CI(
  combined_results_,
  xlimits = c(min(combined_results_$CI_Lower), max(combined_results_$CI_Upper))
) +
  coord_cartesian(xlim = c(1, 3)) +
  labs(caption = "Distribution of 95% CI bounds. Dashed line = True ATE (2.0).")
```

## Point Estimates with Confidence Intervals

The forest plot below shows **mean estimates ± mean CI bounds** for each estimator, ordered by CI width. Narrower CIs with centres close to 2.0 represent the best-performing estimators.

```{r ci_forest_plot, fig.height=4}
plot_est_CI(
  combined_results_,
  custom_title = "Mean ATE Estimates with 95% Confidence Intervals",
  xlimits      = c(min(combined_results_$CI_Lower), max(combined_results_$CI_Upper)),
  true_value   = true_value
) +
  labs(caption = "Dashed line = True ATE (2.0). Ordered by CI width (widest first).")
```

------------------------------------------------------------------------

# Discussion

## Key Findings

The simulation results illustrate several important properties of these estimators in high-dimensional sparse settings:

**Bias:** All four estimators are approximately unbiased when n = 5,000, reflecting their doubly-robust and semiparametrically efficient construction. In smaller samples (n = 1,000), the C-TMLE variants tend to show reduced bias relative to standard TMLE, due to the collaborative regularisation of the propensity score.

**Variance:** The C-TMLE estimators demonstrate a favourable bias-variance trade-off compared to standard TMLE. By selecting a regularised propensity score model collaboratively with the outcome model, C-TMLE avoids overfitting the treatment mechanism — a key source of variance inflation in high-dimensional settings.

**Coverage:** Influence function-based confidence intervals achieve coverage close to the nominal 95% level for all estimators at n = 5,000, validating the inference procedure.

**Practical recommendation:** In high-dimensional sparse settings with large n, C-TMLE1 (Lasso regularisation path) offers the best overall performance, combining computational tractability with strong statistical properties.

## Limitations and Extensions

-   Results are specific to the Toeplitz covariance structure and sparse signal assumed here; performance may differ under denser or differently structured confounding
-   SuperLearner-based nuisance estimation (not implemented here) may further improve performance by data-adaptively combining multiple learners
-   Extension to settings with **missing data** or **non-ignorable missingness** — the focus of the broader PhD research — introduces additional complexity not addressed in this simulation

------------------------------------------------------------------------

# Session Information

```{r session_info}
sessionInfo()
```

------------------------------------------------------------------------

*This simulation study is part of PhD research at the University of Edinburgh, School of Mathematics. For questions or collaboration, contact [asantewaahsarpong\@gmail.com](mailto:asantewaahsarpong@gmail.com).*
